Supplementary Information for\\ ``A weakly structured stem for human origins in Africa''

### 2023-03-15 supplement.tex
abbreviation

However, the replacement of Homo sapiens by Neanderthals in the Levant (e.g.
Kebara 2) complicates the persistence of modern humans in the Arabian Peninsula
after 85ka. The end of MIS 5 coincides with increasing aridity in the region
which would have rendered much the Peninsula uninhabitable, except for a
secluded southern refugia, decreasing the probability of migrations through the
Sinai land bridge \citep{Beyer2021-ym}. The northern route OOA remains a likely
barrier to human migration (as modeled by precipitation amount) until the
Holocene. A southern route across the Red Sea becomes possible during MIS 4
(70ka) until 30ka \citep{Beyer2021-ym}. Even during MIS 4, there were few
plausible routes out of the southern Arabian Peninsula, except along the Gulf
of Oman or gaps in the Arabian Desert through the Zagros Mountains. Neanderthal
remains dating to this period have been identified in the Zagros Mountains,
adding a further potential barrier to expansion
\citep{Yousefi2020-et,Heydari-Guran2021-xy}.

Thus, either an incipient OOA population must have expanded beyond the Near
East by 85ka and then given rise to today’s Eurasian populations or a
successful expansion would likely post-date 65ka. Morphological and
archaeological evidence for Homo sapiens in the Near East reappears by 50ka in
the Levant with the Initial Upper Paleolithic stone tool industry
\citep{Boaretto2021-oi} and a cranium from Manot Cave
\citep{Hershkovitz2015-lq}. There is also strong consensus that a widespread
human presence in Europe and Asia is dated to after 45ka. Archaeological
\citep[e.g.,][]{Fewlass2020-se}, morphological \citep[e.g.,][]{Hublin2020-nk},
and genetic data \citep[e.g.,][]{Prufer2021-ao} all indicate that about 45ka
\emph{Homo sapiens} had expanded into eastern Europe and within 5,000 years
rapidly colonized the entire span of southern Europe. This period further
coincides with the dates for the last Neanderthal presence in Europe at
approximately 40ka \citep{Deviese2017-de,Hublin2017-uy}. Similarly in Asia,
evidence of the Initial Upper Paleolithic occurs 39-45ka as far east as Lake
Baikal. Characteristics of the IUP support a coherent, cultural package which
may have been transmitted by demic expansion \citep{Kuhn2014-kj,Zwyns2021-zv}.
By 45ka, there is additionally an unequivocal \emph{H. sapiens} femur from
southern Siberia which contains derived alleles shared with later Eurasian
populations \citep{Fu2014-am}. The Ust’-Ishim
specimen contains evidence of admixture with Neanderthals, at a level similar
to contemporary East Asian genomes. This supports estimates of Neanderthal
admixture occurring earlier than 45ka, estimated to be by 47-65ka as inferred
from Neanderthal tract lengths \citep{Sankararaman2012-uw}.

>>>>

However, the replacement of Homo sapiens by Neanderthals in the Levant (e.g.
Kebara 2) complicates the persistence of modern humans in the Arabian Peninsula
after 85ka. The end of MIS 5 coincides with increasing aridity in the region
which would have rendered much the Peninsula uninhabitable, except for a
secluded southern refugia, decreasing the probability of migrations through the
Sinai land bridge \citep{Beyer2021-ym}. This barrier to human migration OOA
via the northern route likely remained until the Holocene (as modeled by
precipitation amount).  A southern route across the Red Sea becomes possible
during MIS 4 (70ka) until 30ka \citep{Beyer2021-ym}. Even during MIS 4, there
were few plausible routes out of the southern Arabian Peninsula, except along
the Gulf of Oman or gaps in the Arabian Desert through the Zagros Mountains.
Neanderthal remains dating to this period have been identified in the Zagros
Mountains, adding a further potential barrier to expansion
\citep{Yousefi2020-et,Heydari-Guran2021-xy}.

Thus, either an incipient OOA population must have expanded beyond the Near
East by 85ka and then given rise to today’s Eurasian populations or a
successful expansion would likely post-date 65ka. Morphological and
archaeological evidence for Homo sapiens in the Near East reappears by 50ka in
the Levant with the Initial Upper Paleolithic stone tool industry
\citep{Boaretto2021-oi} and a cranium from Manot Cave
\citep{Hershkovitz2015-lq}. There is also strong consensus that a widespread
human presence in Europe and Asia is dated to after 45ka. Archaeological
\citep[e.g.,][]{Fewlass2020-se}, morphological \citep[e.g.,][]{Hublin2020-nk},
and genetic data \citep[e.g.,][]{Prufer2021-ao} all indicate that about 45ka
\emph{Homo sapiens} had expanded into eastern Europe and within 5,000 years
rapidly colonized the entire span of southern Europe. This period further
coincides with the dates for the last Neanderthal presence in Europe at
approximately 40ka \citep{Deviese2017-de,Hublin2017-uy}. Similarly in Asia,
evidence of the Initial Upper Paleolithic (IUP) occurs 39-45ka as far east as Lake
Baikal. Characteristics of the IUP support a coherent, cultural package which
may have been transmitted by demic expansion \citep{Kuhn2014-kj,Zwyns2021-zv}.
By 45ka, there is additionally an unequivocal \emph{H. sapiens} femur from
southern Siberia which contains derived alleles shared with later Eurasian
populations \citep{Fu2014-am}. The Ust’-Ishim
specimen contains evidence of admixture with Neanderthals, at a level similar
to contemporary East Asian genomes. This supports estimates of Neanderthal
admixture occurring earlier than 45ka, estimated to be by 47-65ka as inferred
from Neanderthal tract lengths \citep{Sankararaman2012-uw}.
